Optimal Finite Element Integration

### 2015-08-10 TOMS_Paper/assembly.tex
Many small fixes

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines for arbitrary multilinear forms. Optimality is defined in terms of floating point operations required to compute a result. We establish a model under which optimality can systematically be reached and comment on the general applicability of this model. The model is integrated in a compiler such that, given a representation of an integration routine, a loop nest in optimal form can automatically be derived. Extensive performance evaluation proves the robustness of the model, showing systematic performance improvements over a number of alternative code generation systems. The effect of low-level optimization is also discussed.
\end{abstract}

@@

As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on aggressive loop optimizations for cache locality, so they are not particularly helpful in our context. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation systems or synthesizers. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. In~\citeN{FFC-TC} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed. It is also worth mentioning the on-going effort to produce a novel form compiler in FEniCS, called UFLACS (\cite{Uflacs}), which adds to the abundant set of code generation (or transformation) systems for assembly operators.

@@

We summarize the contributions of this research as follows
\begin{itemize}
\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration. As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
\item We provide an incremental, constructive proof that optimal implementations can be synthesized by analyzing each monomial appearing in a form
\item We implement the theory in a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is the default optimization system in the Firedrake framework. Our implementation is general: we can apply our results to any forms expressible in Firedrake (and, theoretically, in FEniCS too).
\item We provide an unprecedented (for its extension) performance evaluation comparing several code generation or transformation systems in a range of forms of increasing complexity. This is essential to demonstrate our optimality claim.
\item We comment on the cases in which our characterization of optimality becomes weaker, and intuitions about how we handle them
\item We enrich the set of low-level transformations described in~\citeN{Luporini}
\end{itemize}

\section{Preliminaries}
\label{sec:background}
We review finite element integration and possible implementations using notation and examples adopted in~\cite{quadrature1} and~\cite{Francis}.

We consider the weak formulation of a linear variational problem
\begin{equation}
\begin{split}
Find\ u\ \in U\ such\ that \\
a(u, v) = L(v), \forall v \in V
\end{split}
\end{equation}
where $a$ and $L$ are, respectively, a bilinear and a linear form. The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces. For simplicity, we assume $U = V$ and $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$. The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$. From the solution of the following linear system it is possible to determine a set of coefficients to express $u$
\begin{equation}
A\textbf{u} = b
\end{equation}
in which $A$ and $b$ discretize $a$ and $L$ respectively:
\begin{equation}
\centering
\begin{split}
A_{ij} = a(\phi_i(x), \phi_j(x)) \\
b_i = L(\phi_i(x))
\end{split}
\end{equation}
The matrix $A$ and the vector $b$ are ``assembled'' and subsequently used to solve the linear system through (typically) an iterative method.

@@

\subsection{Quadrature Mode}
Quadrature schemes are conveniently used to numerically evaluate $A_{ij}^K$. For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced. This implies a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated. The numerical integration routine based on quadrature over an element $K$ can be expressed as follows
\begin{equation}
\label{eq:quadrature}
A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
\end{equation}
where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.

@@

These two modes have been implemented in the FEniCS Form Compiler (\cite{FFC-TC}). In this compiler, a heuristic is used to choose the most suitable mode for a given form. It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically (\cite{Fenics}). We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}. For the moment, we just recall that one of the goals of this research is to produce an intelligent system that is capable of reasoning at the monomials level -- not at the form level -- without affecting the code generation cost.

@@

In this section, we provide a characterization of optimality for loop nests of various nature. In order to make the document self-contained, we start with reviewing basic compiler terminology.

@@

% In the following, we should represent loops by their iteration variable.
We find convenient abstracting mathematical properties at the level of loops

@@

Figure~\ref{code:multi_loopnest} shows an example of a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $j$.

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}
\end{CenteredBox}\caption{Multilinear loop nest with sharing}\label{code:multi_loopnest}\end{figure}

@@

For example, by applying the construction described in Proposition~\ref{prop:optimal-mln} to the code in Figure~\ref{code:multi_loopnest}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_optimal}.

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}
\end{CenteredBox}\caption{Multilinear loop nest in optimal form}\label{code:multi_loopnest_optimal}\end{figure}

@@

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
\end{CenteredBox}\caption{Loop nest example}\label{code:loopnest}\end{figure}

The imperfect nest $LN=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear. The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication. We observe that \texttt{F} might contain sub-expressions that depend on a subset of loops only and, as such, hoistable outside of $L_k$ as long as data dependencies are preserved (note that $LN$ is imperfect). One could think of searching for sub-expressions for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count. However, finding or exposing such sub-expressions requires, in general, a full exploration of the expression tree transformation space. This is obviously challenging, and so is the problem of synthesizing an optimal $LN'$ starting from $LN$. Even though we assumed we could generate $LN'$ through a suitable transformation algorithm, the following issues should be addressed
\begin{itemize}
\item as opposed to what happens with hoisting in perfect multilinear loop nests, the temporary variable size would be proportional to the number of non-reduction loops crossed (in the example, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
\item the transformations exposing invariant sub-expressions could require expansion and factorization. In terms of number of operations, the save originating from hoisting could be overwhelmed by the increase in arithmetic complexity within $L_k$ (expansion can increase the operation count, e.g. $A(B+C) = AB + AC$).
\end{itemize}

We shall then refine our definition of optimality for generic loop nests as follows

@@

Note how the definition contemplates the possibility for a nest to have multiple innermost loops. In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations.

@@

In this section, we instantiate Definition~\ref{def:optimality} in our domain of interest, finite element integration. This will require reasoning at two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem, which we reviewed in Section~\ref{sec:background}; and the (partly multilinear) loop nests implementing such forms.

Our point of departure is the example loop nest in Figure~\ref{code:loopnest}. This loop nest is actually a simplified view of a typical bilinear form implementation. $L_e$ represents iteration over the elements of a mesh; $L_i$ derives from using numerical quadrature; $[L_j, L_k]$ implement the computation over test and trial functions. We deliberately omitted useless portions of code to not hinder readability (e.g. matrix insertion) and to avoid tying our discussion to specific forms (e.g. \texttt{F} is unspecified).

@@

\subsection{Memory constraints}
\label{sec:mem-const}
The fact that the iteration space of $L_e$ is so larger than that of other loops suggests we should be cautious when hoisting out of $LN$. Imagine a time stepping loop $L_t$ wraps $LN$. One could think of identifying time-invariant sub-expressions that access both geometry and reference element terms, and pre-evaluate them within $L_t$. Unless adopting complex engineering solutions (e.g. aggressive blocking), which are practically difficult to devise and maintain, this kind of code motion increases the working set by a factor $L$. It is our opinion that the drop in number of operations would be overwhelmed, from the run-time viewpoint, by a much larger memory pressure.

A second, more general observation is that, for certain forms and discretizations, aggressive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). We will provide precise details about this in the following sections. For the moment, and just as an example, note that applying tensor contraction mode (see Section~\ref{sec:background}), which essentially means lifting code outside of $L_e$, requires a set of temporary arrays of size $N\cdot O$; with some discretizations, this can break the local memory threshold.

Based upon these considerations, we add two constraints to $C$ (see Definition~\ref{def:optimality})  
\begin{enumerate}
\item The size of a temporary due to code motion cannot be bigger than the size of the multilinear loop nest iteration space. 
\item The total size of the hoisted temporaries cannot exceed a threshold \texttt{$T_H$}
\end{enumerate}
A corollary of $C_1$ is that hoisting expressions involving geometry terms outside of $L_e$ is now forbidden. Consequently, the search space for optimality becomes smaller.

\subsection{Minimizing the Operation Cost}
Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the sum of the operation cost in the innermost loops is minimum. We now discuss how we can systematically achieve this.

Eliminating sharing from the inner multilinear loops does not suffice. In fact, as suggested in Section~\ref{def:optimality}, we wonder whether, and under what transformations, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.

To answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}. Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}). We interpret this result at the loop nest level: with the input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $LN$, and finally pre-evaluated into temporaries. As part of this pre-evaluation, the reduction induced by $L_i$ vanishes. In the following, we simply refer to this special sort of code hoisting as ``\textit{pre-evaluation}''.

@@

\begin{figure}[h]
\begin{lstlisting}[basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial M:
  $\theta_{w}$ = estimate operation count with pre-evaluation
  $\theta_{wo}$ = estimate operation count without pre-evaluation
  if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
    mark M as candidate for pre-evaluation
for each monomial M:
  if M does not share terms with M$'$, an unmarked monomial:
    extract M into a separate loop nest
    apply pre-evaluation to M
for each expression:
  remove sharing    
\end{lstlisting}
\caption{Intuition of the main algorithm}
\label{code:intuition}
\end{figure}

The point of departure consists of understanding the impact, as number of operations saved or introduced, of pre-evaluation. This is studied ``locally''; that is, for each monomial, in isolation. If we estimate that, for a given monomial, pre-evaluation will decrease the operation count, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}), and the evaluation eventually performed. The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form. Identical tables are mapped to the same temporary. Eventually, sharing is removed from the resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}. The transformed loop nest is as in Figure~\ref{code:loopnest-opt}.

@@

\begin{Prop}
\label{prop:optimal-approach}
Consider an expression comprising a set of monomials $M$. Let $P$ be the set of pre-evaluated monomials, determined as described in Figure~\ref{code:intuition}, and be $Z = M \setminus P$. Assume that:
\begin{enumerate}
\item the cost function employed is optimal; that is, it predicts correctly whether pre-evaluation is profitable or not for a monomial
\item pre-evaluating distinct monomials does not produce identical tables
\item monomials in $P$ do not share terms
\end{enumerate} 
Then, the loop nest $LN$ is optimal under memory constraints $C$, once sharing is removed.
\end{Prop}
\begin{proof}
We first comment on the assumptions. 1) The creation of an optimal cost function is discussed in Section~\ref{sec:op_count}. 2) Identical pre-evaluated tables from distinct monomials could be the result of symmetries in tabulated basis functions. This is however a pathological case. 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms. For simplicity, we do not handle this case, which otherwise would require a ``global'' analysis of the monomials in the form. Further, even though, in practice, assumptions 2) and 3) did not hold, the resulting loop nest would still be nearly-optimal.

@@

B) In absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$, being the cost function optimal due to assumption 1). Assumption 3) guarantees there can be no sharing within $P_Z$, which avoids subtle cases wherein pre-evaluation would result sub-optimal due to destroying sharing-removal opportunities. The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$. This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are ``de-classed'' from $P$ to $Z$: the rationale is that we would have to pay anyway the presence of the shared terms in the innermost loop due to $z$, so adding $p$ to $Z$ does not augment the operation count. 
\end{proof}

@@

We can easily predict $\theta_{wo}$ thanks to our key property, linearity. As explained in Proposition~\ref{prop:optimal-mln}, a simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $MLN_{ops}$. Then, assuming $I$ to be the size of the $L_i$ iteration space, we simply have that $\theta_{wo} = MLN_{ops} \cdot I$.

>>>>

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines for arbitrary multilinear forms. Optimality is defined in terms of floating point operations required to compute a result. We establish a model under which optimality can systematically be reached and comment on the general applicability of this model. The model is integrated in a compiler that, given a representation of an integration routine, derives a loop nest in optimal form. Extensive performance evaluation proves the robustness of the model, showing systematic performance improvements over a number of alternative code generation systems. The effect of low-level optimization is also discussed.
\end{abstract}

@@

As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on loop optimization for cache locality, so they are not particularly helpful in our context. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. In~\citeN{FFC-TC} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed. It is also worth mentioning an on-going effort to produce a novel form compiler, called UFLACS (\cite{Uflacs}), which adds to the already abundant set of code transformation systems for assembly operators.

@@

We summarize the contributions of this research as follows
\begin{itemize}
\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration. As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
\item We provide an incremental, constructive proof that optimal implementations can be obtain by analyzing the monomials appearing in a form
\item We automate the translation of forms into optimal loop nests by a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, the default optimization system in the Firedrake framework.
\item We provide an unprecedented (for its extension) performance evaluation comparing several code transformation systems in a range of forms of increasing complexity. This is essential to demonstrate that our optimality model holds in practice.
\item We comment on the cases in which our characterization of optimality becomes weaker, and provide an intuition of how we handle such forms
\item We enrich the set of low-level transformations described in~\citeN{Luporini}. These make COFFEE a full-fledged system capable of decoupling form compilers from low level code generation.
\end{itemize}

\section{Preliminaries}
\label{sec:background}
We review finite element integration using the same notation and examples adopted in~\citeN{quadrature1} and~\citeN{Francis}.

We consider the weak formulation of a linear variational problem
\begin{equation}
\begin{split}
Find\ u\ \in U\ such\ that \\
a(u, v) = L(v), \forall v \in V
\end{split}
\end{equation}
where $a$ and $L$ are, respectively, a bilinear and a linear form. The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces. For simplicity, we assume $U = V$. Let $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$. The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$. From the solution of the following linear system it is possible to determine a set of coefficients to express $u$
\begin{equation}
A\textbf{u} = b
\end{equation}
in which $A$ and $b$ discretize $a$ and $L$ respectively:
\begin{equation}
\centering
\begin{split}
A_{ij} = a(\phi_i(x), \phi_j(x)) \\
b_i = L(\phi_i(x))
\end{split}
\end{equation}
The matrix $A$ and the vector $b$ are ``assembled'' and subsequently used to solve the linear system through (typically) an iterative method.

@@

\subsection{Quadrature Mode}
Quadrature schemes are typically used to numerically evaluate $A_{ij}^K$. For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced. This implies a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated. The numerical integration routine based on quadrature over an element $K$ can be expressed as follows
\begin{equation}
\label{eq:quadrature}
A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
\end{equation}
where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.

@@

These two modes have been implemented in the FEniCS Form Compiler (\cite{FFC-TC}). In this compiler, a heuristic is used to choose the most suitable mode for a given form. It consists of analysing each monomial in the form, counting the number of derivatives and coefficient functions, and checking if this number is greater than a constant found empirically (\cite{Fenics}). We will later comment on the efficacy of this approach (Section~\ref{sec:optimal-synthesis}. For the moment, we just recall that one of the goals of this research is to produce an intelligent system that goes beyond the dichotomy between quadrature and tensor modes. We will reason in terms of loop nests, code motion, and code pre-evaluation, searching the entire implementation space for an optimal synthesis.

@@

In this section, we provide a characterization of loop nest optimality. In order to make the document self-contained, we start with reviewing basic compiler terminology.

@@

% In the following, we should represent loops by their iteration variable.
It is convenient to map mathematical properties to the loop nest level

@@

Figure~\ref{code:multi_loopnest_a} shows an example of a trivial multilinear loop nest of arity $n=2$ with sharing along dimension $j$.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 7pt \subfigure[With sharing]{\label{code:multi_loopnest_a}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest.code}}}
~~~~~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Optimal form]{\label{code:multi_loopnest_b}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/multilinear_loopnest_opt.code}}}
\end{CenteredBox}\caption{A simple multilinear loop nest}\end{figure}

@@

For example, by applying the construction described in Proposition~\ref{prop:optimal-mln} to the code in Figure~\ref{code:multi_loopnest_a}, we obtain the optimal form in Figure~\ref{code:multi_loopnest_b}.

@@

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest.code}
\end{CenteredBox}\caption{An imperfect loop nest, with one reduction loop ($L_i$), an inner perfect multilinear loop nest ($[L_j, L_k]$), and sharing potentially present in $F$}\label{code:loopnest}\end{figure}

The imperfect nest $LN=[L_e, L_i, L_j, L_k]$ comprises a reduction loop $L_i$ and a perfect doubly nested loop $[L_j, L_k]$, which we assume to be multilinear. The right hand side of the statement computing the multidimensional array \texttt{A} is a generic expression \texttt{F} including standard arithmetic operations such as addition and multiplication, possibly characterized by sharing. Here, one could think of searching for sub-expressions for which the reduction could be pre-evaluated, thus obtaining a decrease proportional to $M$ in the operation count. This is exemplified through Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored} by fixing $F$ in a simpler loop nest.

\begin{figure}[h]\begin{CenteredBox}
{\subfigcapskip = 19pt \subfigure[With reduction]{\label{code:loopnest_red}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_red.code}}}
~~~~~~~~~~~~~~
{\subfigcapskip = 2pt \subfigure[Pre-evaluated reduction]{\label{code:loopnest_nored}\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_nored.code}}}
\end{CenteredBox}\caption{The effect of pre-evaluating reductions. Note the decrease in operation count grows with the loop nest depth.}\end{figure}

It is then obvious that reaching an optimal form $LN'$ is, in the general case, a challenging issue, since a full exploration of the expression tree transformation space is necessary. Even though we assumed we could generate $LN'$ through a suitable transformation algorithm, the following issues should be addressed
\begin{itemize}
\item as opposed to what happens with hoisting in perfect multilinear loop nests, the temporary variable size would be proportional to the number of non-reduction loops crossed (in the example, $N \cdot O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L \cdot N \cdot O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual runtime
\item the transformations exposing sub-expressions amenable to pre-evaluation could increase the arithmetic complexity within $L_k$ (for example, expansion may increase the operation count). This could overwhelm any gain due to pre-evaluation.
\end{itemize}

We shall then refine our definition of optimality for this wider class of loop nests as follows

@@

Note how the definition contemplates the possibility for a nest to have multiple innermost loops. In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations (e.g., see Figure~\ref{code:loopnest_nored}).

@@

In this section, we instantiate Definition~\ref{def:optimality} to our domain of interest, finite element integration. This will require reasoning at two different levels of abstraction: the math, in terms of the multilinear forms arising from the weak variational formulation of a problem; and the (partly multilinear) loop nests implementing such forms.

Our point of departure is the example loop nest in Figure~\ref{code:loopnest}. This loop nest is actually a simplified view of a typical bilinear form implementation. $L_e$ represents iteration over the elements of a mesh; $L_i$ derives from using numerical quadrature; $[L_j, L_k]$ implement the computation over test and trial functions. We deliberately omitted useless portions of code to not hinder readability and to avoid tying our discussion to specific forms.

@@

\subsection{Memory constraints}
\label{sec:mem-const}
The fact that $L >>> M, N, O$ suggests we should be cautious about hoisting out of $LN$. Imagine $LN$ is enclosed in a time stepping loop. One could think of exposing and then hoisting time-invariant sub-expressions. There are several kinds of time-invariant sub-expressions: for example, those depending on the geometry (i.e., depending on $L_e$) and those which do not (e.g., when only involving reference element terms). In the former case, code motion increases the working set by a factor $L$ (unless adopting complex engineering solutions, like blocking non-affine loop nests, which are practically difficult to devise and maintain). The drop in number of operations could then be overwhelmed, from the runtime viewpoint, by a much larger memory pressure.

A second, more general thought is that, for certain forms and discretizations, excessive hoisting can make the working set exceed the size of ``some level of local memory'' (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU). For example, applying tensor contraction mode, which essentially means pre-evaluating geometry-independent expressions outside of $LN$, requires temporary arrays of size $N\cdot O$; with certain discretizations, as detailed in Section~\ref{sec:perf-results-forms}, this can break the local memory threshold.

Based upon these considerations, we add two elements to the set of memory constraints $C$ (see Definition~\ref{def:optimality})  
\begin{enumerate}
\item The size of a temporary due to code motion cannot be larger than that of the multilinear iteration space. 
\item The total amount of memory occupied by the hoisted temporaries cannot exceed a threshold \texttt{$T_H$}
\end{enumerate}
A corollary of $C_1$ is that hoisting expressions involving geometry-dependent terms outside of $LN$, which we discussed at the beginning of this section, is now forbidden. Consequently, the search space for optimality becomes smaller.

\subsection{Minimizing the Operation Cost}
Definition~\ref{def:optimality} states that a necessary condition for a loop nest synthesis to be optimal is that the sum of the innermost loop operation costs is minimum. We now discuss how we can systematically achieve this.

Eliminating sharing from the inner multilinear loops does not suffice. In fact, as suggested in Section~\ref{sec:optimal-impl}, we wonder whether, and under what transformations, the reduction imposed by $L_i$ could be pre-evaluated, thus reducing the operation count.

To answer this question, we make use of a result -- the foundation of tensor contraction mode -- from~\citeN{Kirby:TC}. Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}). We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is then factorized so as to split constant from $[L_i, L_j, L_k]$-dependent terms, the latter ones are hoisted outside of $LN$, and finally pre-evaluated into temporaries. As part of this pre-evaluation, the reduction induced by $L_i$ vanishes. In the following, we simply refer to this special sort of code hoisting as ``\textit{pre-evaluation}''.

@@

\begin{figure}[h]
\begin{lstlisting}[numbers=left, basicstyle=\small\ttfamily, frame=single]
dissect the input expression into monomials
for each monomial M:
  $\theta_{w}$ = estimate operation count with pre-evaluation
  $\theta_{wo}$ = estimate operation count without pre-evaluation
  if $\theta_{w}$ < $\theta_{wo}$ and memory constraints satisfied:
    mark M as candidate for pre-evaluation
for each monomial M:
  if M does not share terms with M$'$, an unmarked monomial:
    extract M into a separate loop nest
    apply pre-evaluation to M
for each expression:
  remove sharing    
\end{lstlisting}
\caption{Intuition of the main algorithm}
\label{code:intuition}
\end{figure}

We study the impact of pre-evaluation, as number of operations saved or introduced, ``locally''; that is, for each monomial, in isolation. If we estimate that, for a given monomial, pre-evaluation will decrease the operation count, then the corresponding sub-expression is extracted, a sequence of transformation steps -- involving expansion, factorization, code motion -- takes place (details in Section~\ref{sec:codegen}), and the evaluation eventually performed. The result is a set of $n$-dimensional tables (these can be seen as ``slices'' of the reference tensor at the math level), $n$ being the arity of the multilinear form. Identical tables are mapped to the same temporary. Eventually, sharing is removed from the resulting expressions by applying a procedure as described in Proposition~\ref{prop:optimal-mln}. The transformed loop nest is as in Figure~\ref{code:loopnest-opt}.

@@

\begin{Prop}
\label{prop:optimal-approach}
Consider an expression comprising a set of monomials $M$. Let $P$ be the set of pre-evaluated monomials, determined as described in Figure~\ref{code:intuition}, and be $Z = M \setminus P$. Assume that:
\begin{enumerate}
\item the cost function $\theta$ employed is optimal; that is, it predicts correctly whether pre-evaluation is profitable or not for a monomial
\item pre-evaluating different monomials does not result in identical tables
\item monomials in $P$ do not share terms
\end{enumerate} 
Then, the loop nest $LN$ is optimal under memory constraints $C$, once sharing is removed.
\end{Prop}
\begin{proof}
We first comment on the assumptions. 1) We postpone the discussion of how to build $\theta$ to Section~\ref{sec:op_count}. 2) Identical pre-evaluated tables from distinct monomials could be the result of symmetries in tabulated basis functions. This is however a pathological case that we can ignore without compromises. 3) In complex forms with several monomials, different pre-evaluation candidates could actually share terms. We abstract from this case, which otherwise would require a ``global'' analysis of the monomials in the form that we believe would not justify the gain (if any) in operation count.

@@

B) In absence of sharing, the statement is trivially true since we would have $LN_{ops_2}' > LN_{ops_2}$, being the cost function optimal due to assumption 1). Assumption 3) guarantees there can be no sharing within $P_Z$, which avoids subtle cases wherein pre-evaluation would be sub-optimal due to destroying sharing-removal opportunities. The last case we have to consider is when $p \in P_Z$ shares at least one term with $z \in Z$. This situation cannot actually occur by construction: all candidates for pre-evaluation sharing terms with monomials in $Z$ are ``de-classed'' from $P$ to $Z$ (see Figure~\ref{code:intuition}, line 8). The rationale is that since we would have to pay anyway the presence of $z$ terms in the innermost loop, adding $p$ to $Z$ does not augment the operation count in our optimality model, so we can just avoid pre-evaluation.

% assunzione 3 è quella che previene la somma di due mass matrix moltiplicate da F0 una e da F1 l'altra...ma non solo ! chesso magari la trial function è derivata in un monomial e nell'altra no, ma la test function è sempre uguale...
\end{proof}

@@

We can easily predict $\theta_{wo}$ thanks to our key property, linearity. This was explained in Proposition~\ref{prop:optimal-mln}. A simple analysis suffices to obtain the cost of a  sharing-free multilinear loop nest, namely $LN_{ops}^{sf}$. Assuming $I$ to be the size of the $L_i$ iteration space, we simply have that $\theta_{wo} = LN_{ops}^{sf} \cdot I$.
